Efficient simulation of individual-based population models

### 2024-03-18 article.qmd
update

Main caption
:::

@@

>>>>

Choices of `CLASS` and `TYPE` arguments for an event creation.
:::

@@

### Intensities

In `IBMPopSim`, the intensity of an event can belong to three classes @sec-event-intensity: individual intensities without interaction between individuals, corresponding to events $e\in\mathcal{E}$, individual intensities with interaction, corresponding to events $e\in\mathcal{E}_W$, and Poisson intensities (homogeneous and inhomogeneous), corresponding to events $e\in\mathcal{P}$.

#### Event creation with individual intensity {.unnumbered}

An event $e\in \mathcal{E}$ (see @eq-no-interaction) has an intensity of the form $\lambda^e(t, I)$ which depends only on the individual `I` and time. Events with such intensity are created using the function
```{R}
#| eval: False
mk_event_individual(type = "TYPE", 
                    name = "NAME",
                    intensity_code = "INTENSITY", ...)
```
The `intensity_code` argument is a character string containing few lines of `C++` code describing the intensity function $\lambda^e(t, I)$. The intensity value has to be stored in a variable called `result` and the available variables for the intensity code are given in @tbl-intensity-variables.

::: {#tbl-intensity-variables}

Variable           Description
------------------ -------------------------------------------------------------
Variable           Description
`I`                Current individual
`J`                Another individual in the population (only for interaction)
`t`                Current time
Model parameters   Depends on the model

`C++` variables available for intensity code
:::

For instance, the intensity code below corresponds to an individual death intensity $\lambda^d(t,I)$ equal to $d_1(a(I,t)) = \alpha_1 \exp(\beta_1 a(I,t))$ for males and $d_2(a(I,t)) = \alpha_2 \exp(\beta_2 a(I,t))$ for females, where $a(I,t)=t-\tau^b$ is the age of the individual $I=(\tau^b, \tau^d,x)$ at time $t$. 
In this case, the intensity function depends on the individuals' age, gender, and on the model parameters $\alpha = (\alpha_1, \alpha_2)$ and $\beta = (\beta_1, \beta_2)$.
```{R}
#| eval: false
death_intensity <- "
    if (I.male) result = alpha_1 * exp(beta_1 * I.age(t));
    else result = alpha_2 * exp(beta_2 * I.age(t));
"
```

#### Event creation with interaction intensity {.unnumbered}

An event $e\in \mathcal{E}_W$ is an event which occurs to an individual at a frequency which is the result of interactions with other members of the population (see @eq-intensity-interaction), and which can be written as $\lambda^e_t(I, Z_t)=\sum_{J\in Z_t} W^e(t, I, J)$ where $W^e(t, I, J)$ is the intensity of the interaction between individual $I$ and individual $J$.

An event $e\in \mathcal{E}_W$ with such intensity is created by calling the function
```{R}
#| eval: false
mk_event_interaction(type = "TYPE",
                     name = "NAME",
                     interaction_code = "INTERACTION_CODE",
                     interaction_type = "random", ...)
```

The `interaction_code` argument contains few lines of `C++` code describing the interaction function $W^e(t, I, J)$. The interaction function value has to be stored in a variable called `result` and the available variables for the intensity code are given in @tbl-intensity-variables. For example, if we set
```{R}
#| eval: false
death_interaction_code <- "result = max(J.size - I.size, 0.);"
```
the death intensity of an individual `I` is the result of the competition between individuals, depending on a characteristic named `size`, as defined in @eq-ex-interaction.

The argument `interaction_type`, set by default at `random`, is the algorithm choice for simulating the model. When `interaction_type=full`, the simulation follows @cnj-no-interaction, `interaction_type=random` it follows @cnj-interaction.
In most cases, the `random` algorithm is much faster than the `full` algorithm, as we illustrate for instance in @sec-example-interaction , where we observe the gain of a factor of 40 between the two algorithms, on a set of standard parameters. This allows in particular to explore parameter sets that give larger population sizes, without reaching computation times that explode.

#### Events creation with Poisson and Inhomogeneous Poisson intensity {.unnumbered}

For events $e\in\mathcal{P}$ with an intensity $\mu^e(t)$ which does not depend on the population, the event intensity is of class `inhomogeneous_poisson` or `poisson` depending on whether or not the intensity depends on time (in the second case the intensity is constant).

For Poisson (constant) intensities the events are created with the function
TODO
